Diverse Genotype-by-Weather Interactions in Switchgrass

### 2024-06-11 manuscript/revision2/revision2.tex
revision2: Coherent M&M between Results, M&M, & SI Appendix

Genotypes from the Gulf and Midwest subpopulations had distinct
phenological timings and distinct patterns of phenological correlations
across our eight common garden sites (Figure~\ref{fig-map}). At the
three Texas common gardens (hereafter `Texas' gardens), located within
the natural range of the Gulf subpopulation, the onset of vegetative
growth for Gulf genotypes (hereafter `green-up') occurred before
Midwestern green-up, and the onset of reproductive growth for Gulf
genotypes (hereafter `flowering') occurred after Midwestern flowering
(Figure~\ref{fig-map} A). At the four northernmost common gardens
(hereafter `North' gardens), located within the natural range of the
Midwest subpopulation, both green-up and flowering of Gulf genotypes
occurred after that of Midwest genotypes. At the Oklahoma common garden,
located near the natural range limits of both the Gulf and the Midwest
subpopulations, Gulf and Midwest green-up occurred over the same period,
and Gulf genotype flowering occurred after Midwestern flowering
(Figure~\ref{fig-map} A). These patterns led to strong negative
phenotypic correlations for the onset of vegetative growth between the
North and Texas gardens, particularly in the Gulf and across all
individuals from the Gulf and Midwest subpopulations (hereafter, `Both'
subpopulations), and contributed to positive phenotypic correlations for
the onset of flowering that had larger magnitudes at more northern
gardens (Figure~\ref{fig-map} B).

@@

\subsubsection{\texorpdfstring{Inference of GxWeather effects for
vegetative and reproductive timing using
\emph{mash}}{Inference of GxWeather effects for vegetative and reproductive timing using mash}}\label{inference-of-gxweather-effects-for-vegetative-and-reproductive-timing-using-mash}

We next looked for evidence of GxWeather by jointly re-estimating
genetic effects of SNPs across all eight common gardens using the
\emph{mash} model, which can flexibly capture many modes of covariance
across gardens (Figure~\ref{fig-map} D; SI Appendix, Sections S1-S5,
Datasets 1-6). We specified three qualitative categories of covariance
structures across gardens: ``canonical'' covariance, with simple
patterns of effect size covariance introduced in the initial \emph{mash}
manuscript; ``data-driven'' covariances derived from common patterns of
SNP effects observed in the data, and ``GxWeather'' covariance,
estimated from the covariance of empirical weather patterns at each
garden at specific times before the phenological event (Table 1; SI
Appendix, Section S1). For example, the more similar that the amount of
rainfall prior to flowering is for genotypes A and B between gardens in
MI and MO, the nearer these covariances will be to one. If rainfall
amounts are uncorrelated, covariance will be near zero; if there is rank
changing (e.g.~high rainfall before flowering of A at MO but not MI, and
high rainfall before flowering for B at MI but not MO), this covariance
will be near negative one. We specified multiple covariance structures
for the canonical, data-driven, and GxWeather categories, each of which
represented a specific pattern of covariance in genetic effect size
estimates that could be included in the model. The phenotypic
correlations for the timing of vegetative growth had moderate negative
correlations between the Texas and North gardens, particularly in the
Gulf subpopulation and Both subpopulations (Figure~\ref{fig-map} B). If
these phenotypic correlations have a genetic component, they could be
partially or completely captured by the covariance structures specified
in the \emph{mash} model.

The GxWeather covariance structures allow hypothesis testing of specific
weather variables as cues for the start of vegetative and reproductive
growth. Say that a SNP in the gene \emph{Flowering Locus C} (FLC), a
well-known flowering time regulator, controls flowering in a
photoperiod-dependent manner. In that case, the joint estimate of
effects for that SNP could have a high mixture proportion, or mass, on a
covariance matrix created using a photoperiod-based environmental cue,
such as day length at some interval prior to flowering. In our data, we
would infer that the effect of that SNP on flowering was caused by a
response to the environmental cue used to construct the GxWeather
covariance structure with the largest mass. Overall, loadings of genetic
effects on these GxWeather matrices in the model, along with the
posterior effect size estimates, provide information on genome-wide
patterns of SNP-environment interaction.

We used a greedy algorithm to select covariance matrices from each
model's set that significantly improved the model likelihood. Five
GxWeather covariance structures were selected, zero to two per
subpopulation: two for green-up date, and three for flowering date. Of
these five, three had mass on them in \emph{mash} models of the strong
effects (Figure~\ref{fig-covar} A-B). Two of these three matrices had
negative covariances between sites between Texas and North gardens
(Figure~\ref{fig-covar} A), while one had all positive or near-zero
covariances. SNP-associated phenotypic effects covaried with different
weather-based cues in the Gulf \& in Both subpopulations
(Figure~\ref{fig-covar} B). In total, 65\% of the posterior weight of
strong SNP effects in the \emph{mash} model of Gulf green-up fell on a
covariance matrix constructed using the covariance of day length 14 days
prior to the date of green-up. The covariance matrix for this weather
cue was visually similar to the observed pattern of phenotypic
correlation for the timing of vegetative growth in the Gulf
subpopulation (Figure~\ref{fig-map} B; Figure~\ref{fig-covar} A).
\emph{Mash} models of the timing of vegetative growth in the Midwest
subpopulation and Both subpopulations did not include this GxWeather
covariance; the Midwest had no weight on any GxWeather covariance, while
Both subpopulations had non-zero weights on two additional GxWeather
covariance types, average temperature one day prior to green-up, and the
day length change in seconds in the day prior to green-up
(Figure~\ref{fig-covar} B). The average temperature covariance matrix
had negative covariances between Texas and North gardens, though not as
strong as the negative phenotypic correlations seen in Both
subpopulations (Figure~\ref{fig-map} B; Figure~\ref{fig-covar} A). Only
the Gulf subpopulation and Both subpopulations had mass on any GxWeather
covariance matrices (Figure~\ref{fig-covar} C) for vegetative growth.

For flowering date, distinct GxWeather covariance structures captured
covariance in effect sizes for SNPs in the Gulf and Midwest
subpopulations. 33\% of SNP effects on flowering in the Gulf
subpopulation covaried with cumulative rainfall in the seven days prior
to flowering (Figure~\ref{fig-covar} D). 22.6\% of SNP effects on
flowering in the Midwest subpopulation covaried with day length change
in the two days prior to flowering (Figure~\ref{fig-covar} D), which had
negative covariances between Texas and North gardens. Neither of these
covariance matrices significantly improved the log-likelihood of the
\emph{mash} model of Both subpopulations, and no GxWeather covariance
matrices had non-zero mass in models of Both subpopulations
(Figure~\ref{fig-covar} E). In five of the six \emph{mash} models of
strong effects, the GxWeather covariance matrices captured a minority of
the posterior weights of the strong effects (Figure~\ref{fig-covar}
C,E); in these five models, the majority of this mass was on various
canonical covariance matrices. These matrices included simple
heterozygosity, with intermediate, positive covariances between all
gardens, and single effect matrices with garden-specific effects (SI
Appendix, Section S1).

We next characterized the pairwise patterns of effects where we were
confident in the sign of the effect at both gardens. If the overall
pattern of phenotypic expression differs between a pair of gardens, this
difference may be comprised of SNP effects that differ in sign between
these gardens (effects with rank-changing GxE between gardens), SNP
effects that differ in magnitude (effects that are large in one garden
or region and smaller in others), and SNP effects that are
indistinguishable in the two gardens (Figure~\ref{fig-effects} A). We
used the local false sign rate (lfsr), an analogue of the lfdr that
establishes confidence in the effect sign, not the effect's difference
from zero, to determine significance. We required lfsr significance (p
\textless{} 0.05) in both gardens to include effects. This means that
our tests for a sign change between gardens carry an equal statistical
burden to those for effects with the same sign.

For green-up date for the Gulf subpopulation, hundreds to thousands of
pairwise effects exhibited a difference in effect sign, between pairs of
Texas and North gardens (Figure~\ref{fig-effects} B; SI Appendix, Fig
S1). 78.7\% of pairwise comparisons between North and Texas gardens had
a difference in sign, while only 28.6\% and 0.2\% of North-North or
Texas-Texas comparisons had a difference in sign, respectively
(Figure~\ref{fig-effects} B). The majority of pairwise effects for
greenup for the Midwest (\textgreater55\%) and Both (\textgreater85\%)
subpopulations were the same sign, and effects most often differed in
magnitude between gardens within and between the regions
(Figure~\ref{fig-effects} B).

For flowering date for the Gulf subpopulation, less than 2\% of pairwise
effects exhibited a difference in effect sign within or between regions
(Figure~\ref{fig-effects} B). More effects differed in magnitude between
the Texas and North regions than within these regions (42.7\% vs
\textless20\%; Figure~\ref{fig-effects} B), while the majority of
effects had no GxE. The Midwest population had relatively few
significant effects for flowering, but a large proportion of these
differed in sign between Texas and North regions (42.7\%) or within the
North region (65.4\%). Finally, in Both subpopulations, less than 20\%
of pairwise effects differed in sign (Figure~\ref{fig-effects} B). Most
differences in sign were between TX1, the southernmost garden, and all
other gardens (SI Appendix, Figure S1). Similarly, more effects that
differed in magnitude included gardens in the Texas region
(52.3-55.9\%), and most effect pairs in the North region were not
distinguishable (91.5\%).

@@

We sought additional experimental support the occurrence and location of
our significant SNP effect re-estimates using an independent pseudo-F2
mapping population created from Gulf \& Midwest individuals and grown at
the same sites (Figure~\ref{fig-qtl} A,B). We conducted quantitative
trait loci (QTL) mapping of flowering as functions of five environmental
cues that we also used as covariance matrices in \emph{mash}, and
identified eight QTL for flowering date, eight QTL for flowering as a
function of day length change two days prior, one QTL for the start of
vegetative growth, and two QTL for vegetative growth as a function of
daylength change one day prior, all of which showed QTL by environment
interactions (SI Appendix, Figure S2). All QTL for flowering overlapped
one or more homologs from rice or \emph{A. thaliana} with functionally
validated roles in flowering (SI Appendix, Dataset 7). All flowering and
green-up QTL intervals contained at least one SNP significant in at
least one \emph{mash} run at a log10-transformed Bayes Factor
\textgreater{} 2, or in the 1\% tail of significance, whichever was
stricter (SI Appendix, Dataset 8). We also looked for enrichments of
\emph{mash} SNPs in the 1\% tail of significance (the `\emph{mash} 1\%
tail') within each QTL interval. At the 5\% level, three QTL had
enrichments of SNPs in the \emph{mash} 1\% tail. Overall, there were
five significant enrichments (p \textless{} 0.05, hypergeometric test)
of SNPs in the \emph{mash} 1\% tail in the QTL intervals. Thus, we were
able to experimentally support some of our re-estimates of SNP effects
with a QTL mapping experiment using a separate mapping population.

@@

Our analysis of the timing of vegetative growth in the Gulf and in Both
subpopulations revealed many alleles with sign changes in their effects
between the Texas and North gardens (Figure~\ref{fig-effects} B). As
phenological timings are major components of plant fitness, this result
supports theoretical models that local adaptation should involve
antagonistic pleiotropy at the level of individual loci (23--26); it
expands the genetics of local adaptation research to more than two field
sites and to a wider range of genetic variation by using effects from
GWAS in common gardens located throughout the species' range to
determine that rank-changing GxE for phenological timings is common in
small genomic regions (12, 38, 39).

@@

Whenever possible, plant material will be shared upon request. Source
data and code to replicate these analyses are available at:
https://github.com/Alice-MacQueen/pvdiv-phenology-gxe.git. SNP data to
replicate these analyses are available from the UT dataverse at
https://doi.org/link.

@@

In 2019, we scored two phenological events every two days in two mapping
populations of switchgrass, a diversity panel and a pseudo-F2 cross,
planted at eight common garden locations (35, 39, 40). We scored the
onset of vegetative growth, or green-up date, as the day of the year
when 50\% of the tiller area of the crown of the plant cut the previous
year had green growth. The onset of reproductive growth, or flowering
date, was the day of the year when 50\% of the plant tillers had
panicles undergoing anthesis.

The formation and resequencing of the diversity panel has been described
previously (33). The diversity panel contained 134 sequenced, clonally
propagated individuals from the Midwest genetic subpopulation, and 229
from the Gulf genetic subpopulation. To allow for the possibility that
different subpopulations had different strengths of connection between
our phenotypes and genotypes (41), we conducted three sets of genetic
analyses: on Gulf and Midwest genotypes separately, and on both
subpopulations together (`Both' subpopulations). Analyses to determine
narrow-sense heritability (h\textsuperscript{2}) for green-up and
flowering were done using linear mixed models and followed (33) (SI
Appendix, Section S6).

To confirm candidate genomic regions and patterns of allelic effects
found in the diversity panel, we analyzed flowering in an outbred
pseudo-F2 cross between four individuals, two Midwest and two Gulf
individuals. The formation of this mapping population has been described
previously (35); additional details on QTL mapping can be found in SI
Appendix, Section S7. To be directly comparable to the diversity panel
data, only 2019 phenology data from the pseudo-F2 cross from the same
eight common garden sites were used.

@@

To use \emph{mash} on our diversity panel, we had to specify both a
relatively uncorrelated set of covariance matrices, which in our case
defined types of GxE and GxWeather between gardens, and we had to
specify subsets of SNP effect estimates and standard errors for our
traits at each common garden. To specify a set of covariance matrices,
we first defined many covariance matrices, including GxWeather matrices
that represented the correlation in weather cues between gardenes before
the phenological event (SI Appendix, Section S1), then implemented a
model selection approach that used a greedy algorithm to evaluate if the
log likelihood of the \emph{mash} model was significantly improved as
additional covariance matrices were included (SI Appendix, Section S4).
To specify subsets of SNP effect estimates and standard errors to use
for both the greedy algorithm and with the optimal set of covariance
matrices, we first calculated best linear unbiased predictors (BLUPs)
for each phenological trait in each genetic subpopulation and each
common garden (SI Appendix, Section S2). Next, we determined effect
estimates for 8.8 to 12.3 million SNPs per subpopulation by conducting
garden-specific GWAS on these BLUPs using \(k\) vectors of singular
values to correct for population structure, where \(k\) was the smallest
integer value that made the genomic control coefficient closest to 1 (SI
Appendix, Section S2). Singular values were computed using singular
value decomposition of the matrix of all SNPs, with iterative SNP
pruning and removal of regions in long-range linkage disequilibrium
(42). Third, to make the \emph{mash} models computationally feasible, we
extracted two subsets of SNP effect estimates and standard errors from
our GWAS effect estimates: 1) effects from a subset of ``strong'' tests
corresponding to stronger effects on our traits; 2) results from a
random subset of all tests to correspond to an unbiased representation
of all effects (SI Appendix, Section S3). We used the subset of random
effects in our greedy \emph{mash} algorithm (SI Appendix, Section S4)
and used the subset of strong effects in our \emph{mash} models with the
optimal set of covariance matrices (SI Appendix, Section S5).

@@

Using the lfsr rather than the local false discovery rate (lfdr) is a
critical change in our ability to detect alleles directly contributing
to rank-changing GxE between environments. The lfdr, like other measures
of FDR, focuses on if we have enough evidence to reject the null
hypothesis that an effect j is 0, or that there is a significant effect.
Previous studies of antagonistic pleiotropy (e.g. (39)) have used the
lfdr or equivalent statistical tests to detect antagonistic pleiotropy.
These tests were conservative, in that they required two non-zero
effects of different signs, while tests for differential sensitivity
required only one non-zero effect. This previous work recognized that
this testing bias could lead to undercounting occurrences of
antagonistic pleiotropy (27, 29), and sought to reduce it by permutation
(28). However, using the lfsr to test for allelic effects that differ in
sign does not undercount these occurences, as this statistic answers a
fundamentally different question. For each effect \(j\), the \(lfsr_j\)
is defined as the probability that we make an error in the sign of
effect \(j\) if we were forced to declare the effect positive or
negative (37). Thus, rather than asking ``Are these two effects
different?'' - as we reasonably expect two effects to be, even if this
difference cannot be measured - the local false sign rate answers a more
meaningful question: Can we be confident in the sign of this effect?

@@

\bibitem[\citeproctext]{ref-savolainen2013ecological}
\CSLLeftMargin{38. }%
\CSLRightInline{O. Savolainen, M. Lascoux, J. Merilä, Ecological
genomics of local adaptation. \emph{Nature Reviews Genetics}
\textbf{14}, 807--820 (2013).}

\bibitem[\citeproctext]{ref-lowry2019qtl}
\CSLLeftMargin{39. }%
\CSLRightInline{D. B. Lowry, \emph{et al.}, QTL\(\times\) environment
interactions underlie adaptive divergence in switchgrass across a large
latitudinal gradient. \emph{Proceedings of the National Academy of
Sciences} \textbf{116}, 12933--12941 (2019).}

\bibitem[\citeproctext]{ref-lovell2021}
\CSLLeftMargin{40. }%
\CSLRightInline{J. T. Lovell, \emph{et al.},
\href{https://doi.org/10.1038/s41586-020-03127-1}{Genomic mechanisms of
climate adaptation in polyploid bioenergy switchgrass}. \emph{Nature},
1--7 (2021).}

\bibitem[\citeproctext]{ref-kortefarlow2013}
\CSLLeftMargin{41. }%
\CSLRightInline{A. Korte, A. Farlow, The advantages and limitations of
trait analysis with GWAS: A review. \emph{Plant methods} \textbf{9}, 29
(2013).}

\bibitem[\citeproctext]{ref-privuxe92017}
\CSLLeftMargin{42. }%
\CSLRightInline{F. Privé, H. Aschard, M. G. B. Blum,
\href{http://dx.doi.org/10.1101/190926}{Efficient management and
analysis of large-scale genome-wide data with two r packages: Bigstatsr
and bigsnpr} (2017).}

>>>>

Genotypes from the Gulf and Midwest subpopulations had distinct
phenological timings and distinct patterns of phenological correlations
across our eight common garden sites (Figure~\ref{fig-map}). At the
three Texas common gardens (hereafter `Texas' gardens), located within
the natural range of the Gulf subpopulation, the onset of vegetative
growth, or `green-up date', for Gulf genotypes occurred before
Midwestern green-up, and the onset of reproductive growth, or `flowering
date', for Gulf genotypes occurred after Midwestern flowering
(Figure~\ref{fig-map} A). At the four northernmost common gardens
(hereafter `North' gardens), located within the natural range of the
Midwest subpopulation, both green-up and flowering of Gulf genotypes
occurred after that of Midwest genotypes. At the Oklahoma common garden,
located near the natural range limits of both the Gulf and the Midwest
subpopulations, Gulf and Midwest green-up occurred over the same period,
and Gulf genotype flowering occurred after Midwestern flowering
(Figure~\ref{fig-map} A). These patterns led to strong negative
phenotypic correlations for the onset of vegetative growth between the
North and Texas gardens, particularly in the Gulf and across all
individuals from the Gulf and Midwest subpopulations (hereafter, `Both'
subpopulations), and contributed to positive phenotypic correlations for
the onset of flowering that had larger magnitudes at more northern
gardens (Figure~\ref{fig-map} B).

@@

\subsubsection{Inference of genome-wide patterns of GxE and GxWeather
for vegetative and reproductive
timing}\label{inference-of-genome-wide-patterns-of-gxe-and-gxweather-for-vegetative-and-reproductive-timing}

We expected patterns of GxE and GxWeather to vary at the locus,
subpopulation, and trait level; thus we were interested in modeling the
different ways effects might covary across our eight common gardens. We
jointly re-estimated genetic effects of a subset of SNPs across all
eight common gardens using \emph{mash} models, which can flexibly
capture many modes of covariance across gardens (Figure~\ref{fig-map} D;
SI Appendix, Sections S1-S5, Datasets 1-6). To obtain genetic effects,
we used subsets of effects from garden- and subpopulation GWAS (SI
Appendix, Section S2-S3; Table S1). To obtain covariance structures, we
specified both covariance structures introduced in the initial
\emph{mash} manuscript, and GxWeather covariance matrices estimated from
the covariance of empirical weather patterns at each garden at specific
times before the phenological event (Table 1; SI Appendix, Section S1).
Then, we used a greedy algorithm to select covariance matrices from each
model's set that significantly improved the model likelihood (SI
Appendix, Section S4; Table S2). Finally, we fit \emph{mash} models
using the selected set of covariance matrices on a subset relatively
unlinked (r\textsuperscript{2} \textless{} 0.2), `strong' genetic
effects with low \emph{p}-values (SI Appendix, Section S5).

The loadings of genetic effects onto the GxE and GxWeather covariance
matrices included in each model provide information on genome-wide
patterns of SNP-environment interaction. For example, the phenotypic
correlations for the onset of vegetative growth had moderate negative
correlations between the Texas and North gardens, particularly in the
Gulf subpopulation and Both subpopulations (Figure~\ref{fig-map} B). If
these phenotypic correlations have a genetic component, they could be
partially or completely captured by the covariance structures specified
in the \emph{mash} model. In this case, the joint estimate of effects
for many SNPs could have high mixture proportions, or mass, on a
covariance matrix that captures the negative correlation between Texas
and North gardens. In our data, daylength 14 days prior to green-up date
has negative correlations between Texas and North gardens
(Figure~\ref{fig-covar} A); if many SNP effects have mass on this
GxWeather matrix, we would infer that the effects of those SNPs have
that form of GxWeather covariation.

Eight GxWeather covariance structures were selected by the greedy
algorithm, one to two per subpopulation: five for green-up date, and
three for flowering date (SI Appendix, Table S2). Of these eight, five
had mass on them in \emph{mash} models of the strong effects
(Figure~\ref{fig-covar} A-B). Three of these five matrices had negative
covariances between sites between Texas and North gardens
(Figure~\ref{fig-covar} A), while two had all positive or near-zero
covariances.

For green-up date, SNP-associated phenotypic effects covaried with
different weather-based cues in the Gulf \& in Both subpopulations
(Figure~\ref{fig-covar} B). In total, 65\% of the posterior weight of
strong SNP effects in the \emph{mash} model of Gulf green-up fell on a
covariance matrix constructed using the covariance of day length 14 days
prior to the date of green-up. The covariance matrix for this weather
cue was visually similar to the observed pattern of phenotypic
correlation for the timing of vegetative growth in the Gulf
subpopulation (Figure~\ref{fig-map} B; Figure~\ref{fig-covar} A).
\emph{Mash} models of the timing of vegetative growth in the Midwest
subpopulation and Both subpopulations did not include this GxWeather
covariance; the Midwest had no weight on any GxWeather covariance, while
Both subpopulations had non-zero weights on two additional GxWeather
covariance types, average temperature one day prior to green-up, and the
day length change in seconds in the day prior to green-up
(Figure~\ref{fig-covar} B). The average temperature covariance matrix
had negative covariances between Texas and North gardens, though not as
strong as the negative phenotypic correlations seen in Both
subpopulations (Figure~\ref{fig-map} B; Figure~\ref{fig-covar} A). Only
the Gulf subpopulation and Both subpopulations had mass on any GxWeather
covariance matrices (Figure~\ref{fig-covar} C) for vegetative growth.

For flowering date, distinct GxWeather covariance structures captured
covariance in effect sizes for SNPs in the Gulf and Midwest
subpopulations. 33\% of SNP effects on flowering in the Gulf
subpopulation covaried with cumulative rainfall in the seven days prior
to flowering (Figure~\ref{fig-covar} D). 22.6\% of SNP effects on
flowering in the Midwest subpopulation covaried with day length change
in the two days prior to flowering (Figure~\ref{fig-covar} D), which had
negative covariances between Texas and North gardens. Neither of these
covariance matrices significantly improved the log-likelihood of the
\emph{mash} model of Both subpopulations, and no GxWeather covariance
matrices had non-zero mass in models of Both subpopulations
(Figure~\ref{fig-covar} E). In five of the six \emph{mash} models of
strong effects, the GxWeather covariance matrices captured a minority of
the posterior weights of the strong effects (Figure~\ref{fig-covar}
C,E); in these five models, the majority of this mass was on various
canonical covariance matrices. These matrices included simple
heterozygosity, with intermediate, positive covariances between all
gardens, and single effect matrices with garden-specific effects.

\subsubsection{Frequency of rank-changing GxE in significant SNP
effects}\label{frequency-of-rank-changing-gxe-in-significant-snp-effects}

We expected to observe common rank-changing GxE at the level of
individual loci as this is a key theoretical prediction of local
adaptation (23--26). To determine the frequency of rank-changing GxE, we
used the local false sign rate (lfsr), an analogue of the lfdr that
establishes confidence in the effect sign, not the effect's difference
from zero, to determine significance. We required lfsr significance (p
\textless{} 0.05) in both gardens to include effects. This means that
our tests for a sign change between gardens carry an equal statistical
burden to those for effects with the same sign. We separated kinds of
effects at the level of individual loci into SNP effects that differ in
sign between gardens (effects with rank-changing GxE between gardens),
SNP effects that differ in magnitude (effects that are large in one
garden or region and smaller in others), and SNP effects that are
indistinguishable in two gardens (Figure~\ref{fig-effects} A).

For green-up date for the Gulf subpopulation, hundreds to thousands of
pairwise effects exhibited rank-changing GxE, or a difference in sign
between pairs of Texas and North gardens (Figure~\ref{fig-effects} B; SI
Appendix, Fig S1). 78.7\% of pairwise comparisons between North and
Texas gardens had a difference in sign, while only 28.6\% and 0.2\% of
North-North or Texas-Texas comparisons had a difference in sign,
respectively (Figure~\ref{fig-effects} B). The majority of pairwise
effects for greenup for the Midwest (\textgreater55\%) and Both
(\textgreater85\%) subpopulations were the same sign, and effects most
often differed in magnitude between gardens within and between the
regions (Figure~\ref{fig-effects} B).

For flowering date for the Gulf subpopulation, less than 2\% of pairwise
effects exhibited rank-changing GxE, or a difference in effect sign
within or between regions (Figure~\ref{fig-effects} B). More effects
differed in magnitude between the Texas and North regions than within
these regions (42.7\% vs \textless20\%; Figure~\ref{fig-effects} B),
while the majority of effects had no GxE. The Midwest population had
relatively few significant effects for flowering, but a large proportion
of these differed in sign between Texas and North regions (42.7\%) or
within the North region (65.4\%). Finally, in Both subpopulations, less
than 20\% of pairwise effects differed in sign (Figure~\ref{fig-effects}
B). Most differences in sign were between TX1, the southernmost garden,
and all other gardens (SI Appendix, Figure S1). Similarly, more effects
that differed in magnitude included gardens in the Texas region
(52.3-55.9\%), and most effect pairs in the North region were not
distinguishable (91.5\%).

@@

We sought additional experimental support the occurrence and location of
our significant SNP effect re-estimates using an independent pseudo-F2
mapping population created from Gulf \& Midwest individuals and grown at
the same sites (Figure~\ref{fig-qtl} A,B). We conducted quantitative
trait loci (QTL) mapping of flowering as functions of five environmental
cues that we also used as covariance matrices in \emph{mash}, and
identified eight QTL for flowering date, eight QTL for flowering as a
function of day length change two days prior, one QTL for the start of
vegetative growth, and two QTL for vegetative growth as a function of
daylength change one day prior, all of which showed QTL by environment
interactions (SI Appendix, Figure S2). All QTL for flowering overlapped
one or more homologs from rice or \emph{A. thaliana} with functionally
validated roles in flowering (SI Appendix, Dataset 7); the QTL on Chr04K
overlapped a gene recently functionally validated for flowering in
switchgrass (38). All flowering and green-up QTL intervals contained at
least one SNP significant in at least one \emph{mash} run at a
log10-transformed Bayes Factor \textgreater{} 2, or in the 1\% tail of
significance, whichever was stricter (SI Appendix, Dataset 8). We also
looked for enrichments of \emph{mash} SNPs in the 1\% tail of
significance (the `\emph{mash} 1\% tail') within each QTL interval. At
the 5\% level, three QTL had enrichments of SNPs in the \emph{mash} 1\%
tail. Overall, there were five significant enrichments (p \textless{}
0.05, hypergeometric test) of SNPs in the \emph{mash} 1\% tail in the
QTL intervals. Thus, we were able to experimentally support the genomic
windows of some re-estimates of significant SNP effects with a QTL
mapping experiment using a separate mapping population.

@@

Our analysis of the timing of vegetative growth in the Gulf and in Both
subpopulations revealed many alleles with rank-changing GxE, or sign
changes in their effects between the Texas and North gardens
(Figure~\ref{fig-effects} B). As phenological timings are major
components of plant fitness, this result supports theoretical models
that local adaptation should involve antagonistic pleiotropy at the
level of individual loci (23--26). In addition, it expands the genetics
of local adaptation research to more than two field sites and to a wider
range of genetic variation by re-estimating effects from GWAS to
determine that rank-changing GxE for phenological timings is common in
small genomic regions (12, 39, 40).

@@

Whenever possible, plant material will be shared upon request. Source
data and code to replicate these analyses are available at:
https://github.com/Alice-MacQueen/pvdiv-phenology-gxe.git. SNP data to
replicate these analyses are available from the UT dataverse at
\url{https://doi.org/10.18738/T8/A604BU}.

@@

In 2019, we scored two phenological events every two days in two mapping
populations of switchgrass, a diversity panel and a pseudo-F2 cross,
planted at eight common garden locations (35, 40, 41). We scored the
onset of vegetative growth, or green-up date, as the day of the year
when 50\% of the tiller area of the crown of the plant cut the previous
year had green growth. The onset of reproductive growth, or flowering
date, was the day of the year when 50\% of the plant tillers had
panicles undergoing anthesis.

The formation and resequencing of the diversity panel has been described
previously (33). The diversity panel contained 134 sequenced, clonally
propagated individuals from the Midwest genetic subpopulation, and 229
from the Gulf genetic subpopulation. To allow for the possibility that
different subpopulations had different strengths of connection between
our phenotypes and genotypes (42), we conducted three sets of genetic
analyses: on Gulf and Midwest genotypes separately, and on both
subpopulations together (`Both' subpopulations). Analyses to determine
narrow-sense heritability (h\textsuperscript{2}) for green-up and
flowering were done using linear mixed models and followed (33) (SI
Appendix, Section S6).

To confirm candidate genomic regions found using \emph{mash} on the
diversity panel, we analyzed flowering in an outbred pseudo-F2 cross
between four individuals, two Midwest and two Gulf individuals. The
formation of this mapping population has been described previously (35);
additional details on QTL mapping can be found in SI Appendix, Section
S7. To be directly comparable to the diversity panel data, only 2019
phenology data from the pseudo-F2 cross from the same eight common
garden sites were used.

@@

To use \emph{mash} on our diversity panel, we had to specify both a
relatively uncorrelated set of covariance matrices, which in our case
defined types of GxE and GxWeather between gardens, and we had to
specify subsets of SNP effect estimates and standard errors for our
traits at each common garden. To specify a set of covariance matrices,
we first defined many covariance matrices, including GxWeather matrices
that represented the correlation in weather cues between gardenes before
the phenological event (SI Appendix, Section S1), then implemented a
model selection approach that used a greedy algorithm to evaluate if the
log likelihood of the \emph{mash} model was significantly improved as
additional covariance matrices were included (SI Appendix, Section S4).
To specify subsets of SNP effect estimates and standard errors to use
for both the greedy algorithm and with the optimal set of covariance
matrices, we first calculated best linear unbiased predictors (BLUPs)
for each phenological trait in each genetic subpopulation and each
common garden (SI Appendix, Section S2). Next, we determined effect
estimates for 8.8 to 12.3 million SNPs per subpopulation by conducting
garden-specific GWAS on these BLUPs using \(k\) vectors of singular
values to correct for population structure, where \(k\) was the smallest
integer value that made the genomic control coefficient closest to 1 (SI
Appendix, Section S2). Singular values were computed using singular
value decomposition of the matrix of all SNPs, with iterative SNP
pruning and removal of regions in long-range linkage disequilibrium
(43). Third, to make the \emph{mash} models computationally feasible, we
extracted two subsets of SNP effect estimates and standard errors from
our GWAS effect estimates: (i) effects from a subset of ``strong'' tests
corresponding to stronger effects on our traits; (ii) results from a
random subset of all tests to correspond to an unbiased representation
of all effects (SI Appendix, Section S3). We used the subset of random
effects in our greedy \emph{mash} algorithm (SI Appendix, Section S4)
and used the subset of strong effects in our \emph{mash} models with the
optimal set of covariance matrices (SI Appendix, Section S5).

@@

Using the lfsr rather than the local false discovery rate (lfdr) is a
critical change in our ability to detect alleles directly contributing
to rank-changing GxE between environments. The lfdr, like other measures
of FDR, focuses on if we have enough evidence to reject the null
hypothesis that an effect j is 0, or that there is a significant effect.
Previous studies of antagonistic pleiotropy (e.g. (40)) have used the
lfdr or equivalent statistical tests to detect antagonistic pleiotropy.
These tests were conservative, in that they required two non-zero
effects of different signs, while tests for differential sensitivity
required only one non-zero effect. This previous work recognized that
this testing bias could lead to undercounting occurrences of
antagonistic pleiotropy (27, 29), and sought to reduce it by permutation
(28). However, using the lfsr to test for allelic effects that differ in
sign does not undercount these occurences, as this statistic answers a
fundamentally different question. For each effect \(j\), the \(lfsr_j\)
is defined as the probability that we make an error in the sign of
effect \(j\) if we were forced to declare the effect positive or
negative (37). Thus, rather than asking ``Are these two effects
different?'' - as we reasonably expect two effects to be, even if this
difference cannot be measured - the local false sign rate answers a more
meaningful question: Can we be confident in the sign of this effect?

@@

\bibitem[\citeproctext]{ref-choi2023}
\CSLLeftMargin{38. }%
\CSLRightInline{S. Choi, \emph{et al.},
\href{https://doi.org/10.1093/jxb/erad255}{A single amino acid change
led to structural and functional differentiation of {\emph{PvHd1}} to
control flowering in switchgrass}. \emph{Journal of Experimental Botany}
\textbf{74}, 5532--5546 (2023).}

\bibitem[\citeproctext]{ref-savolainen2013ecological}
\CSLLeftMargin{39. }%
\CSLRightInline{O. Savolainen, M. Lascoux, J. Merilä, Ecological
genomics of local adaptation. \emph{Nature Reviews Genetics}
\textbf{14}, 807--820 (2013).}

\bibitem[\citeproctext]{ref-lowry2019qtl}
\CSLLeftMargin{40. }%
\CSLRightInline{D. B. Lowry, \emph{et al.}, QTL\(\times\) environment
interactions underlie adaptive divergence in switchgrass across a large
latitudinal gradient. \emph{Proceedings of the National Academy of
Sciences} \textbf{116}, 12933--12941 (2019).}

\bibitem[\citeproctext]{ref-lovell2021}
\CSLLeftMargin{41. }%
\CSLRightInline{J. T. Lovell, \emph{et al.},
\href{https://doi.org/10.1038/s41586-020-03127-1}{Genomic mechanisms of
climate adaptation in polyploid bioenergy switchgrass}. \emph{Nature},
1--7 (2021).}

\bibitem[\citeproctext]{ref-kortefarlow2013}
\CSLLeftMargin{42. }%
\CSLRightInline{A. Korte, A. Farlow, The advantages and limitations of
trait analysis with GWAS: A review. \emph{Plant methods} \textbf{9}, 29
(2013).}

\bibitem[\citeproctext]{ref-privuxe92017}
\CSLLeftMargin{43. }%
\CSLRightInline{F. Privé, H. Aschard, M. G. B. Blum,
\href{http://dx.doi.org/10.1101/190926}{Efficient management and
analysis of large-scale genome-wide data with two r packages: Bigstatsr
and bigsnpr} (2017).}
